---
title: "GradBoostR Vignette"
author: "Nic Coxen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    keep_md: true
knitr:
  opts_chunk:
    cache: false

vignette: >
  %\VignetteIndexEntry{Introduction to GradBoostR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(tidyr)
opts_chunk$set(fig.width=9, fig.height=6, fig.path="Figures/", out.width="100%")
opts_chunk$set(comment="", fig.align="center", fig.retina=2)
options(rmarkdown.html_vignette.check_title = FALSE)
options(knitr.kable.NA = '')
library(GradBoostR)
set.seed(123)
```

## Example Dataset

```{r example data}
X <- matrix(rnorm(300), ncol = 3)
y_reg <- X[,1] * 2 + rnorm(100)
y_class <- factor(ifelse(X[,1] + X[,2] > 0, "A", "B"))

df <- data.frame(
  X1 = X[,1],
  X2 = X[,2],
  X3 = X[,3],
  y_reg = y_reg,
  y_class = y_class
)
```

```{r}
head(df,8)
```

# Visualizing the dataset

```{r data-visual}
df %>% ggplot() +
  aes(X1, y_reg) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(
    title = "Example Dataset: X1 vs y_reg",
    x = "X1",
    y = "Regression Target"
  )
```

# Random Forest Regression

```{r random forest regression}
m_rf <- gradboostr_fit(X, y_reg, method = "rf")
pred_rf <- gradboostr_predict(m_rf, X)
head(pred_rf,8)
```
# Prediction vs Truth

```{r rf-reg-plot}
df_rf <- data.frame(True = y_reg, Pred = pred_rf)

df_rf %>% ggplot() +
  aes(True, Pred) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_abline(color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Random Forest Regression: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
  
```

# Random Forest Classification

```{r random forest classification}
m_rf_c <- gradboostr_fit(X, y_class, method = "rf_class")
pred_class <- gradboostr_predict(m_rf_c, X, type = "response")
head(pred_class,8)
```
# Classification Boundary

```{r rf-class-boundary}
df_class <- data.frame(
  X1 = X[,1],
  X2 = X[,2],
  Class = pred_class
)

df_class %>% ggplot() +
  aes(X1, X2, color = Class) +
  geom_point(alpha = 0.6) +
  theme_grey() +
  labs(
    title = "Random Forest Classification Boundary",
    x = "X1",
    y = "X2",
    color = "Predicted Class" 
    )
```

```{r rf-class-bar}

df_class %>% ggplot() +
  aes(Class) +
  geom_bar(fill = "violet" ) +
  theme_classic() +
  labs(
    title = "Random Forest Classification: Class Distribution",
    x = "Class",
    y = "Count"
  )
```

## Gradient Boosting Regression

```{r gbm regression}
m_gbm <- gradboostr_fit(X, y_reg, method = "gbm", learning_rate = 0.1, n_trees = 50)
pred_gbm <- gradboostr_predict(m_gbm, X)
head(pred_gbm,8)
```

# Prediction vs Truth

```{r gbm-reg-plot}
df_gbm <- data.frame(True = y_reg, Pred = pred_gbm)

df_gbm %>% ggplot()  + 
  aes(True, Pred) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_abline(color = "darkcyan", linetype = "dotted", linewidth = 1.35) +
  theme_light() +
  labs(
    title = "GBM Regression: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
```

# Neural Network Regression

```{r neural net regression}
m_nn <- gradboostr_fit(X, y_reg, method = "nn", hidden = 10, epochs = 50)
pred_nn <- gradboostr_predict(m_nn, X)
head(pred_nn,8)
```
# Prediction vs Truth

```{r nn-reg-plot}
df_nn <- data.frame(True = y_reg, Pred = pred_nn)

df_nn %>%
  ggplot() + aes(True, Pred) +
  geom_point(alpha = 0.6) +
  geom_jitter(color = "dodgerblue") + 
  geom_abline(color = "red", linetype = "dashed") +
   theme_grey() +
   labs(
    title = "Neural Network Regression: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
```

# NN vs RF vs GBM Comparison

```{r combined-model-plot}
df_all <- data.frame(
  True = y_reg,
  RF = pred_rf,
  GBM = pred_gbm,
  NN = pred_nn
)

library(tidyr)

df_long <- pivot_longer(df_all, cols = c("RF", "GBM", "NN"),
                        names_to = "Model", values_to = "Pred")

df_long %>% ggplot() +
  aes(True, Pred, color = Model) +
  geom_point(alpha = 0.5) +
  geom_abline(color = "black", linetype = "dashed") +
  theme_gray() +
  labs(
    title = "Model Comparison: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
```

# NN Residual Plot

```{r nn-residuals}
df_res <- data.frame(
  Pred = pred_nn,
  Resid = y_reg - pred_nn
)

df_res %>% ggplot() + 
  aes(Pred, Resid) +
  geom_point(alpha = 0.6, color = "purple") +
  theme_minimal() +
  labs(
    title = "NN Residuals (hidden = 10)",
    x = "Predicted",
    y = "Residual"
  )
```

# Tuning Hidden Layer Size

```{r nn-tuning-hidden}
sizes <- c(5, 10, 20, 40)

results <- lapply(sizes, function(h) {
  m <- gradboostr_fit(X, y_reg, method = "nn", hidden = h, epochs = 50)
  pred <- gradboostr_predict(m, X)
  rmse <- sqrt(mean((pred - y_reg)^2))
  data.frame(hidden = h, RMSE = rmse)
})

do.call(rbind, results)
```

# Visualizing Hidden Layer Tuning

```{r nn-tuning-plot}
df_tune <- do.call(rbind, results)

df_tune %>% ggplot() + 
  aes(hidden, RMSE) +
  geom_line(color = "darkgreen") +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Effect of Hidden Layer Size on NN Performance",
    x = "Hidden Units",
    y = "RMSE"
  )
```

# Training Time Benchmark

```{r benchmark}


bench_gbm <- system.time({
  gbm_model <- gradboostr_fit(X, y_reg, method = "gbm",
                              learning_rate = 0.1, n_trees = 50)
})

bench_rf <- system.time({
  rf_model <- gradboostr_fit(X, y_reg, method = "rf")
})

bench_nn <- system.time({
  nn_model <- gradboostr_fit(X, y_reg, method = "nn",
                             hidden = 10, epochs = 50)
})

bench_gbm
bench_rf
bench_nn
```

# Benchmark plot

```{r benchmark-plot}
df_bench <- data.frame(
  Model = c("GBM", "RF", "NN"),
  Time  = c(
    bench_gbm["elapsed"],
    bench_rf["elapsed"],
    bench_nn["elapsed"]
  )
)

df_bench %>% ggplot() + 
  aes(Model, Time) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Training Time Comparison",
    y = "Elapsed Time (seconds)"
  )
```

# NN Example with Scaled Inputs

```{r nn-vs-others}
X_scaled <- scale(X)

m_nn_scaled <- gradboostr_fit(X_scaled, y_reg, method = "nn", hidden = 10, epochs = 50)
pred_nn_scaled <- gradboostr_predict(m_nn_scaled, X_scaled)

head(pred_nn_scaled,8)
```
# Learning Rate Example

```{r gbm-learning-rate}
# Example data
X <- matrix(rnorm(300), ncol = 3)
y_reg <- X[,1] * 2 + rnorm(100)

# Fit models with different learning rates
m_lr_01 <- gradboostr_fit(X, y_reg, method = "gbm",
                          learning_rate = 0.1, n_trees = 50)

m_lr_005 <- gradboostr_fit(X, y_reg, method = "gbm",
                           learning_rate = 0.05, n_trees = 50)

m_lr_001 <- gradboostr_fit(X, y_reg, method = "gbm",
                           learning_rate = 0.01, n_trees = 50)

# Predictions
pred_01  <- gradboostr_predict(m_lr_01,  X)
pred_005 <- gradboostr_predict(m_lr_005, X)
pred_001 <- gradboostr_predict(m_lr_001, X)

# RMSE helper
rmse <- function(a, b) sqrt(mean((a - b)^2))

# Results table
df_lr <- data.frame(
  learning_rate = c(0.1, 0.05, 0.01),
  RMSE = c(
    rmse(pred_01,  y_reg),
    rmse(pred_005, y_reg),
    rmse(pred_001, y_reg)
  )
)

df_lr
```

```{r gbm-learning-rate-plot}
df_lr %>% ggplot() +
  aes(learning_rate, RMSE) +
  geom_point(size = 3, color = "darkcyan") +
  geom_line(color = "violet") +
  theme_minimal() +
  labs(
    title = "Effect of Learning Rate on GBM Performance",
    x = "Learning Rate",
    y = "RMSE"
  )
```

# Neural Network Learning Rate

```{r nn-learning-rate}
# Example dataset
set.seed(1)
X <- matrix(rnorm(300), ncol = 3)
y_reg <- X[,1] * 2 + rnorm(100)

# Scale inputs (NNs perform better with scaled features)
X_scaled <- scale(X)

# Fit NN models with different learning rates
m_lr_01  <- gradboostr_fit(X_scaled, y_reg, method = "nn",
                           hidden = 10, epochs = 80, learning_rate = 0.1)

m_lr_005 <- gradboostr_fit(X_scaled, y_reg, method = "nn",
                           hidden = 10, epochs = 80, learning_rate = 0.05)

m_lr_001 <- gradboostr_fit(X_scaled, y_reg, method = "nn",
                           hidden = 10, epochs = 80, learning_rate = 0.01)
# Predictions
pred_01  <- gradboostr_predict(m_lr_01,  X_scaled)
pred_005 <- gradboostr_predict(m_lr_005, X_scaled)
pred_001 <- gradboostr_predict(m_lr_001, X_scaled)

# RMSE helper
rmse <- function(a, b) sqrt(mean((a - b)^2))

# Results table
df_lr <- data.frame(
  learning_rate = c(0.1, 0.05, 0.01),
  RMSE = c(
    rmse(pred_01,  y_reg),
    rmse(pred_005, y_reg),
    rmse(pred_001, y_reg)
  )
)

df_lr
```

```{r nn-learning-rate-plot}
df_lr %>% ggplot() +
  aes(learning_rate, RMSE) +
  geom_point(size = 3, color = "darkorange") +
  geom_line(color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Effect of Learning Rate on Neural Network Performance",
    x = "Learning Rate",
    y = "RMSE"
  )
```

# Effect of Epochs

```{r nn-epochs}
epochs_grid <- c(40, 80, 160)

res_epochs <- lapply(epochs_grid, function(e) {
  m <- gradboostr_fit(
    X_scaled, y_reg,
    method        = "nn",
    hidden        = 10,
    epochs        = e,
    learning_rate = 0.05
  )
  p <- gradboostr_predict(m, X_scaled)
  data.frame(epochs = e, RMSE = rmse(p, y_reg))
})

df_epochs <- do.call(rbind, res_epochs)
df_epochs
```
```{r nn-epochs-plot}
df_epochs %>% ggplot() +
  aes(epochs, RMSE) +
  geom_point(size = 3, color = "darkorange") +
  geom_line(color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Effect of Epochs on NN Performance",
    x = "Epochs",
    y = "RMSE"
  )
```

# Grid search

```{r nn-grid}
hidden_grid <- c(5, 10, 20)
epochs_grid <- c(40, 80)
lr_grid     <- c(0.1, 0.05, 0.01)

grid <- expand.grid(
  hidden        = hidden_grid,
  epochs        = epochs_grid,
  learning_rate = lr_grid
)

res_grid <- lapply(seq_len(nrow(grid)), function(i) {
  h  <- grid$hidden[i]
  e  <- grid$epochs[i]
  lr <- grid$learning_rate[i]

  m <- gradboostr_fit(
    X_scaled, y_reg,
    method        = "nn",
    hidden        = h,
    epochs        = e,
    learning_rate = lr
  )
  p <- gradboostr_predict(m, X_scaled)

  data.frame(
    hidden        = h,
    epochs        = e,
    learning_rate = lr,
    RMSE          = rmse(p, y_reg)
  )
})

df_grid <- do.call(rbind, res_grid)
df_grid
```

```{r Best-grid}
df_grid[which.min(df_grid$RMSE), ]
```

```{r nn-grid-heatmap}
df_grid %>% ggplot() +
  aes(hidden, epochs, fill = RMSE) +
  geom_tile(color = "grey70") +
  facet_wrap(~ learning_rate) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(
    title = "NN Grid Search: Hidden, Epochs, Learning Rate",
    x = "Hidden Units",
    y = "Epochs",
    fill = "RMSE"
  )
```
# Fit Stump

### Create a synthetic regression dataset with 3 features designed to highlight differences between simple and complex models. 

```{r data}
set.seed(683)

X <- cbind(
  x1 = rnorm(300),
  x2 = rnorm(300),
  x3 = rnorm(300) 
)

y <- ifelse(X[,1] > 0, 3 + X[,2], -3 + X[,2]) + rnorm(300, sd = 0.4)

df <- data.frame(X1 = X[,1], X2 = X[,2], X3 = X[,3], y = y)
```

```{r data_snippet}
head(df,8)
```





# Fit stump (decision stump)

#### A decision stump is the simplest possible regression tree: it makes one split on one feature, producing two constant predictions.

#### fit_stump() implements this by exhaustively scanning:

#### every feature in the input matrix X, and

#### every possible threshold between sorted feature values,

#### and for each candidate split it computes the squared‑error loss of predicting:

#### the mean of the left partition for all points left of the threshold, and

#### the mean of the right partition for all points right of the threshold.

#### The function returns the split (feature + threshold) that yields the lowest total squared error, along with the two predicted values for the left and right regions.

```{r fit_stump}
stump <- fit_stump(X, y)
stump
```

```{r predict_stump}
stump_predict <- function(X, stump) {
  f  <- stump$feature + 1  
  thr <- stump$threshold
  lv  <- stump$left_val
  rv  <- stump$right_val
  
  ifelse(X[, f] <= thr, lv, rv)
}
```

```{r stump_fit}
stump <- fit_stump(X, y)
pred_stump <- stump_predict(X, stump)

head(pred_stump, 20)
```

### Generating predictions for other models

```{r gbm_model}
m_rf <- gradboostr_fit(X, y, method = "rf")
pred_rf <- gradboostr_predict(m_rf, X)

```


```{r rf_model}
m_gbm <- gradboostr_fit(
  X, y,
  method        = "gbm",
  learning_rate = 0.1,
  n_trees       = 50
)
pred_gbm <- gradboostr_predict(m_gbm, X)


```


# RMSE    
### RMSE is computed for each model to determine accuracy.

```{r models_rmse}
rmse <- function(a, b) sqrt(mean((a - b)^2))

data.frame(
  Model = c("Stump", "Random Forest", "GBM"),
  RMSE  = c(
    rmse(pred_stump, y),
    rmse(pred_rf,    y),
    rmse(pred_gbm,   y)
  )
)


```
## Comparison visual

```{r visual_rmse}

df_compare <- data.frame(
  Truth = y,
  Stump = pred_stump,
  RF    = pred_rf,
  GBM   = pred_gbm
)

df_long <- pivot_longer(
  df_compare,
  cols = c("Stump", "RF", "GBM"),
  names_to = "Model",
  values_to = "Pred"
)

df_long %>% ggplot() + 
  aes(Truth, Pred) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~ Model) +
  theme_minimal() +
  labs(
    title = "Prediction vs Truth for Stump, Random Forest, and GBM",
    subtitle = "Perfect predictions lie on the dashed diagonal line",
    x = "True Values",
    y = "Predicted Values"
  )

```

## Visuals for synthetic regression dataset.

```{r visual_data1}

df %>% ggplot() + 
  aes(X1, y, color = X2) +
  geom_point(alpha = 0.7) +
  scale_color_gradient2(low = "blue", high = "red", mid = "white") +
  theme_minimal() +
  labs(
    title = "Synthetic Regression Data",
    subtitle = "Two regions defined by the sign of X1",
    x = "X1",
    y = "Target y",
    color = "X2"
  )

```

```{r visual_data2}
df %>% ggplot() + 
  aes(X1, X2, color = y) +
  geom_point(alpha = 0.8) +
  scale_color_gradient2(low = "blue", high = "red", mid = "white") +
  theme_minimal() +
  labs(
    title = "Nonlinear Structure of the Synthetic Dataset",
    subtitle = "Clear boundary at X1 = 0",
    x = "X1",
    y = "X2",
    color = "y"
  )

```
# Neural Network with `nn_train_arma`

#### The GradBoostR package includes a lightweight neural‑network implementation written in C++ using RcppArmadillo.
#### This module provides a fast, minimal, fully transparent single‑hidden‑layer neural network suitable for regression tasks.
#### It is designed to be easy to inspect, modify, and extend, making it ideal for educational purposes and small‑scale modeling.


```{r}
m_nn <- nn_train_arma(X, y, hidden_units = 10, epochs = 100, learning_rate = 0.01)
```


```{r structure}
str(m_nn)
```
## Returned object guide:

#### $W1   # p × h weight matrix for hidden layer
#### $b1   # h × 1 bias vector for hidden layer
#### $W2   # h × 1 weight matrix for output layer
#### $b2   # scalar bias for output layer


## Predict using        `nn_predict_model`

```{r}
pred_nn <- nn_predict_model(X, m_nn)

head(pred_nn)
```
```{r nn_example}
set.seed(123)

# Synthetic nonlinear dataset
X <- cbind(
  x1 = rnorm(600),
  x2 = rnorm(600)
)

y <- sin(X[,1]) + X[,2]^2 + rnorm(600, sd = 0.1)

# Train NN
m_nn <- nn_train_arma(X, y, hidden_units = 10, epochs = 100, learning_rate = 0.01)

# Predict
pred_nn <- nn_predict_model(X, m_nn)


head(pred_nn)

```

# Model performance

```{r}
rmse <- function(a, b) sqrt(mean((a - b)^2))
rmse(pred_nn, y)

```

```{r predictions vs truth}

df_plot <- data.frame(
  Truth = y,
  Pred  = pred_nn
)

df_plot %>% ggplot() + 
  aes(Truth, Pred) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_abline(color = "blue", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Neural Network Predictions vs Truth",
    subtitle = "Single-hidden-layer MLP trained via nn_train_arma()",
    x = "True Values",
    y = "Predicted Values"
  )

```

# Fitting a GBM Model `gbm_predict()`

```{r parameters}
learning_rate <- 0.1
n_trees <- 20

```


```{r residuals}
residuals <- y
stumps <- vector("list", n_trees)

```


```{r boosting}
for (m in seq_len(n_trees)) {
  st <- fit_stump(X, residuals)
  stumps[[m]] <- st
  
  pred_m <- ifelse(
    X[, st$feature + 1] <= st$threshold,
    st$left_val,
    st$right_val
  )
  
  residuals <- residuals - learning_rate * pred_m
}

```


```{r initial value}
init_value <- mean(y)

```


```{r predict}
pred_gbm <- gbm_predict(X, stumps, learning_rate, init_value)

```


```{r evaluate}
rmse <- function(a, b) sqrt(mean((a - b)^2))
rmse(pred_gbm, y)

```

```{r gbm visual}
df_plot <- data.frame(
  Truth = y,
  Pred  = pred_gbm
)

df_plot %>% ggplot() + 
  aes(Truth, Pred) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_abline(color = "red", linetype = "dashed") +
  theme_bw() +
  labs(
    title = "GBM Predictions vs Truth",
    subtitle = "Gradient boosting with decision stumps",
    x = "True Values",
    y = "Predicted Values"
  )
```

## Decision surface `gbm_predict()`

```{r grid}
# Create a grid of points
x1_seq <- seq(min(X[,1]), max(X[,1]), length.out = 150)
x2_seq <- seq(min(X[,2]), max(X[,2]), length.out = 150)

grid <- expand.grid(
  x1 = x1_seq,
  x2 = x2_seq
)

X_grid <- as.matrix(grid)
head(X_grid,8)
```

```{r predict grid}
grid$Pred <- gbm_predict(
  X_grid,
  stumps,
  learning_rate,
  init_value
)

```


```{r visual grid}

grid %>% ggplot() + 
  aes(x1, x2, fill = Pred) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = mean(grid$Pred)
  ) +
  theme_minimal() +
  labs(
    title = "GBM Decision Surface",
    subtitle = "Gradient boosting with decision stumps",
    x = "X1",
    y = "X2",
    fill = "Prediction"
  )

```

## Random forest with `rf_predict_fast_class()`

```{r synthetic classification data}
set.seed(783)

X <- cbind(
  x1 = rnorm(300),
  x2 = rnorm(300)
)

# Nonlinear classification boundary
y <- ifelse(X[,1] + sin(2 * X[,2]) > 0, 1, 0)
```


```{r factor}
y_factor <- factor(y)
```


```{r train}
forest <- rf_class_fit_fast(
  X,
  y,
  n_trees   = 50,
  max_depth = 5,
  mtry      = 1,
  use_gini  = TRUE
)
```

```{r predict_forest}
pred_class <- rf_predict_fast_class(X, forest)
```

```{r convert}
pred_factor <- factor(pred_class)
```

## Probabilities `rf_predict_fast_class_prob()`

```{r probabilities}
pred_prob <- rf_predict_fast_class_prob(X, forest, n_classes = 2)
```

### Accuracy

```{r evaluate accuracy}
mean(pred_factor == y_factor)
```
## Decision boundry `rf_predict_fast_class()`

```{r visual decisi0n boundry}

# Build grid
x1_seq <- seq(min(X[,1]), max(X[,1]), length.out = 200)
x2_seq <- seq(min(X[,2]), max(X[,2]), length.out = 200)

grid <- expand.grid(x1 = x1_seq, x2 = x2_seq)
grid_mat <- as.matrix(grid)

# Predict on grid
grid$Pred <- rf_predict_fast_class(grid_mat, forest)

# Prepare point data
df_points <- data.frame(
  x1 = X[,1],
  x2 = X[,2],
  y_factor = y_factor
)

# Plot
grid %>% ggplot() + 
  aes(x1, x2, fill = Pred) +
  geom_tile(alpha = 0.8) +
  scale_fill_gradient(low = "blue", high = "red") +
  geom_point(
    data = df_points,
    aes(x1, x2, color = y_factor),
    inherit.aes = FALSE,
    size = 1.2
  ) +
  theme_minimal() +
  labs(
    title = "Random Forest Decision Boundary",
    subtitle = "rf_predict_fast_class()",
    x = "X1",
    y = "X2",
    fill = "Predicted Class",
    color = "True Class"
  )
```

