---
title: "GradBoostR Vignette"
author: "Nic Coxen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    keep_md: true
knitr:
  opts_chunk:
    cache: false

vignette: >
  %\VignetteIndexEntry{Introduction to GradBoostR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(tidyr)
opts_chunk$set(fig.width=9, fig.height=6, fig.path="Figures/", out.width="100%")
opts_chunk$set(comment="", fig.align="center", fig.retina=2)
options(rmarkdown.html_vignette.check_title = FALSE)
options(knitr.kable.NA = '')
library(GradBoostR)
set.seed(123)
```

## Example Dataset

```{r example data}
X <- matrix(rnorm(300), ncol = 3)
y_reg <- X[,1] * 2 + rnorm(100)
y_class <- factor(ifelse(X[,1] + X[,2] > 0, "A", "B"))

df <- data.frame(
  X1 = X[,1],
  X2 = X[,2],
  X3 = X[,3],
  y_reg = y_reg,
  y_class = y_class
)
```

```{r}
head(df,8)
```

# Visualizing the dataset

```{r data-visual}
df %>% ggplot() +
  aes(X1, y_reg) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(
    title = "Example Dataset: X1 vs y_reg",
    x = "X1",
    y = "Regression Target"
  )
```

# Random Forest Regression

```{r random forest regression}
m_rf <- gradboostr_fit(X, y_reg, method = "rf")
pred_rf <- gradboostr_predict(m_rf, X)
head(pred_rf,8)
```
# Prediction vs Truth

```{r rf-reg-plot}
df_rf <- data.frame(True = y_reg, Pred = pred_rf)

df_rf %>% ggplot() +
  aes(True, Pred) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_abline(color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Random Forest Regression: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
  
```

# Random Forest Classification

```{r random forest classification}
m_rf_c <- gradboostr_fit(X, y_class, method = "rf_class")
pred_class <- gradboostr_predict(m_rf_c, X, type = "response")
head(pred_class,8)
```
# Classification Boundary

```{r rf-class-boundary}
df_class <- data.frame(
  X1 = X[,1],
  X2 = X[,2],
  Class = pred_class
)

df_class %>% ggplot() +
  aes(X1, X2, color = Class) +
  geom_point(alpha = 0.6) +
  theme_grey() +
  labs(
    title = "Random Forest Classification Boundary",
    x = "X1",
    y = "X2",
    color = "Predicted Class" 
    )
```

```{r rf-class-bar}

df_class %>% ggplot() +
  aes(Class) +
  geom_bar(fill = "darkcyan" , alpha = 0.7) +
  theme_classic() +
  labs(
    title = "Random Forest Classification: Class Distribution",
    x = "Class",
    y = "Count"
  )
```

## Gradient Boosting Regression

```{r gbm regression}
m_gbm <- gradboostr_fit(X, y_reg, method = "gbm", learning_rate = 0.1, n_trees = 50)
pred_gbm <- gradboostr_predict(m_gbm, X)
head(pred_gbm,8)
```

# Prediction vs Truth

```{r gbm-reg-plot}
df_gbm <- data.frame(True = y_reg, Pred = pred_gbm)

ggplot(df_gbm, aes(True, Pred)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_abline(color = "black", linetype = "dotted") +
  theme_light() +
  labs(
    title = "GBM Regression: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
```

# Neural Network Regression

```{r neural net regression}
m_nn <- gradboostr_fit(X, y_reg, method = "nn", hidden = 10, epochs = 50)
pred_nn <- gradboostr_predict(m_nn, X)
head(pred_nn,8)
```
# Prediction vs Truth

```{r nn-reg-plot}
df_nn <- data.frame(True = y_reg, Pred = pred_nn)

df_nn %>%
  ggplot() + aes(True, Pred) +
  geom_point(alpha = 0.6) +
  geom_jitter(color = "dodgerblue") + 
  geom_abline(color = "red", linetype = "dashed") +
   theme_grey() +
   labs(
    title = "Neural Network Regression: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
```

# NN vs RF vs GBM Comparison

```{r combined-model-plot}
df_all <- data.frame(
  True = y_reg,
  RF = pred_rf,
  GBM = pred_gbm,
  NN = pred_nn
)

library(tidyr)

df_long <- pivot_longer(df_all, cols = c("RF", "GBM", "NN"),
                        names_to = "Model", values_to = "Pred")

df_long %>% ggplot() +
  aes(True, Pred, color = Model) +
  geom_point(alpha = 0.5) +
  geom_abline(color = "black", linetype = "dashed") +
  theme_gray() +
  labs(
    title = "Model Comparison: Prediction vs Truth",
    x = "True Values",
    y = "Predicted Values"
  )
```

# NN Residual Plot

```{r nn-residuals}
df_res <- data.frame(
  Pred = pred_nn,
  Resid = y_reg - pred_nn
)

ggplot(df_res, aes(Pred, Resid)) +
  geom_point(alpha = 0.6, color = "purple") +
  theme_minimal() +
  labs(
    title = "NN Residuals (hidden = 10)",
    x = "Predicted",
    y = "Residual"
  )
```

# Tuning Hidden Layer Size

```{r nn-tuning-hidden}
sizes <- c(5, 10, 20, 40)

results <- lapply(sizes, function(h) {
  m <- gradboostr_fit(X, y_reg, method = "nn", hidden = h, epochs = 50)
  pred <- gradboostr_predict(m, X)
  rmse <- sqrt(mean((pred - y_reg)^2))
  data.frame(hidden = h, RMSE = rmse)
})

do.call(rbind, results)
```

# Visualizing Hidden Layer Tuning

```{r nn-tuning-plot}
df_tune <- do.call(rbind, results)

ggplot(df_tune, aes(hidden, RMSE)) +
  geom_line(color = "darkgreen") +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Effect of Hidden Layer Size on NN Performance",
    x = "Hidden Units",
    y = "RMSE"
  )
```

# Training Time Benchmark

```{r benchmark}


bench_gbm <- system.time({
  gbm_model <- gradboostr_fit(X, y_reg, method = "gbm",
                              learning_rate = 0.1, n_trees = 50)
})

bench_rf <- system.time({
  rf_model <- gradboostr_fit(X, y_reg, method = "rf")
})

bench_nn <- system.time({
  nn_model <- gradboostr_fit(X, y_reg, method = "nn",
                             hidden = 10, epochs = 50)
})

bench_gbm
bench_rf
bench_nn
```

# Benchmark plot

```{r benchmark-plot, fig.width=12, fig.height=9}
df_bench <- data.frame(
  Model = c("GBM", "RF", "NN"),
  Time  = c(
    bench_gbm["elapsed"],
    bench_rf["elapsed"],
    bench_nn["elapsed"]
  )
)

ggplot(df_bench, aes(Model, Time)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Training Time Comparison",
    y = "Elapsed Time (seconds)"
  )
```

# NN Example with Scaled Inputs

```{r nn-vs-others}
X_scaled <- scale(X)

m_nn_scaled <- gradboostr_fit(X_scaled, y_reg, method = "nn", hidden = 10, epochs = 50)
pred_nn_scaled <- gradboostr_predict(m_nn_scaled, X_scaled)

head(pred_nn_scaled,8)
```
# Learning Rate Example

```{r gbm-learning-rate}
# Example data
X <- matrix(rnorm(300), ncol = 3)
y_reg <- X[,1] * 2 + rnorm(100)

# Fit models with different learning rates
m_lr_01 <- gradboostr_fit(X, y_reg, method = "gbm",
                          learning_rate = 0.1, n_trees = 50)

m_lr_005 <- gradboostr_fit(X, y_reg, method = "gbm",
                           learning_rate = 0.05, n_trees = 50)

m_lr_001 <- gradboostr_fit(X, y_reg, method = "gbm",
                           learning_rate = 0.01, n_trees = 50)

# Predictions
pred_01  <- gradboostr_predict(m_lr_01,  X)
pred_005 <- gradboostr_predict(m_lr_005, X)
pred_001 <- gradboostr_predict(m_lr_001, X)

# RMSE helper
rmse <- function(a, b) sqrt(mean((a - b)^2))

# Results table
df_lr <- data.frame(
  learning_rate = c(0.1, 0.05, 0.01),
  RMSE = c(
    rmse(pred_01,  y_reg),
    rmse(pred_005, y_reg),
    rmse(pred_001, y_reg)
  )
)

df_lr
```

```{r gbm-learning-rate-plot}
df_lr %>% ggplot() +
  aes(learning_rate, RMSE) +
  geom_point(size = 3, color = "darkcyan") +
  geom_line(color = "violet") +
  theme_minimal() +
  labs(
    title = "Effect of Learning Rate on GBM Performance",
    x = "Learning Rate",
    y = "RMSE"
  )
```

# Neural Network Learning Rate

```{r nn-learning-rate}
# Example dataset
set.seed(1)
X <- matrix(rnorm(300), ncol = 3)
y_reg <- X[,1] * 2 + rnorm(100)

# Scale inputs (NNs perform better with scaled features)
X_scaled <- scale(X)

# Fit NN models with different learning rates
m_lr_01  <- gradboostr_fit(X_scaled, y_reg, method = "nn",
                           hidden = 10, epochs = 80, learning_rate = 0.1)

m_lr_005 <- gradboostr_fit(X_scaled, y_reg, method = "nn",
                           hidden = 10, epochs = 80, learning_rate = 0.05)

m_lr_001 <- gradboostr_fit(X_scaled, y_reg, method = "nn",
                           hidden = 10, epochs = 80, learning_rate = 0.01)
# Predictions
pred_01  <- gradboostr_predict(m_lr_01,  X_scaled)
pred_005 <- gradboostr_predict(m_lr_005, X_scaled)
pred_001 <- gradboostr_predict(m_lr_001, X_scaled)

# RMSE helper
rmse <- function(a, b) sqrt(mean((a - b)^2))

# Results table
df_lr <- data.frame(
  learning_rate = c(0.1, 0.05, 0.01),
  RMSE = c(
    rmse(pred_01,  y_reg),
    rmse(pred_005, y_reg),
    rmse(pred_001, y_reg)
  )
)

df_lr
```

```{r nn-learning-rate-plot}
df_lr %>% ggplot() +
  aes(learning_rate, RMSE) +
  geom_point(size = 3, color = "darkorange") +
  geom_line(color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Effect of Learning Rate on Neural Network Performance",
    x = "Learning Rate",
    y = "RMSE"
  )
```

# Effect of Epochs

```{r nn-epochs}
epochs_grid <- c(40, 80, 160)

res_epochs <- lapply(epochs_grid, function(e) {
  m <- gradboostr_fit(
    X_scaled, y_reg,
    method        = "nn",
    hidden        = 10,
    epochs        = e,
    learning_rate = 0.05
  )
  p <- gradboostr_predict(m, X_scaled)
  data.frame(epochs = e, RMSE = rmse(p, y_reg))
})

df_epochs <- do.call(rbind, res_epochs)
df_epochs
```
```{r nn-epochs-plot}
df_epochs %>% ggplot() +
  aes(epochs, RMSE) +
  geom_point(size = 3, color = "darkorange") +
  geom_line(color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Effect of Epochs on NN Performance",
    x = "Epochs",
    y = "RMSE"
  )
```

# Grid search

```{r nn-grid}
hidden_grid <- c(5, 10, 20)
epochs_grid <- c(40, 80)
lr_grid     <- c(0.1, 0.05, 0.01)

grid <- expand.grid(
  hidden        = hidden_grid,
  epochs        = epochs_grid,
  learning_rate = lr_grid
)

res_grid <- lapply(seq_len(nrow(grid)), function(i) {
  h  <- grid$hidden[i]
  e  <- grid$epochs[i]
  lr <- grid$learning_rate[i]

  m <- gradboostr_fit(
    X_scaled, y_reg,
    method        = "nn",
    hidden        = h,
    epochs        = e,
    learning_rate = lr
  )
  p <- gradboostr_predict(m, X_scaled)

  data.frame(
    hidden        = h,
    epochs        = e,
    learning_rate = lr,
    RMSE          = rmse(p, y_reg)
  )
})

df_grid <- do.call(rbind, res_grid)
df_grid
```

```{r Best-grid}
df_grid[which.min(df_grid$RMSE), ]
```

```{r nn-grid-heatmap}
df_grid %>% ggplot() +
  aes(hidden, epochs, fill = RMSE) +
  geom_tile(color = "grey70") +
  facet_wrap(~ learning_rate) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(
    title = "NN Grid Search: Hidden, Epochs, Learning Rate",
    x = "Hidden Units",
    y = "Epochs",
    fill = "RMSE"
  )
```
# Fit Stump

### Create a synthetic regression dataset with 3 features designed to highlight differences between simple and complex models. 

```{r data}
set.seed(683)

X <- cbind(
  x1 = rnorm(300),
  x2 = rnorm(300),
  x3 = rnorm(300) 
)

y <- ifelse(X[,1] > 0, 3 + X[,2], -3 + X[,2]) + rnorm(300, sd = 0.4)

df <- data.frame(X1 = X[,1], X2 = X[,2], X3 = X[,3], y = y)
```

```{r data_snippet}
head(df,8)
```





# Fit stump (decision stump)

#### A decision stump is the simplest possible regression tree: it makes one split on one feature, producing two constant predictions.

#### fit_stump() implements this by exhaustively scanning:

#### every feature in the input matrix X, and

#### every possible threshold between sorted feature values,

#### and for each candidate split it computes the squared‑error loss of predicting:

#### the mean of the left partition for all points left of the threshold, and

#### the mean of the right partition for all points right of the threshold.

#### The function returns the split (feature + threshold) that yields the lowest total squared error, along with the two predicted values for the left and right regions.

```{r fit_stump}
stump <- fit_stump(X, y)
stump
```

```{r predict_stump}
stump_predict <- function(X, stump) {
  f  <- stump$feature + 1  
  thr <- stump$threshold
  lv  <- stump$left_val
  rv  <- stump$right_val
  
  ifelse(X[, f] <= thr, lv, rv)
}
```

```{r stump_fit}
stump <- fit_stump(X, y)
pred_stump <- stump_predict(X, stump)

head(pred_stump, 20)
```

### Generating predictions for other models

```{r gbm_model}
m_rf <- gradboostr_fit(X, y, method = "rf")
pred_rf <- gradboostr_predict(m_rf, X)

```


```{r rf_model}
m_gbm <- gradboostr_fit(
  X, y,
  method        = "gbm",
  learning_rate = 0.1,
  n_trees       = 50
)
pred_gbm <- gradboostr_predict(m_gbm, X)


```


# RMSE    
### RMSE is computed for each model to determine accuracy.

```{r models_rmse}
rmse <- function(a, b) sqrt(mean((a - b)^2))

data.frame(
  Model = c("Stump", "Random Forest", "GBM"),
  RMSE  = c(
    rmse(pred_stump, y),
    rmse(pred_rf,    y),
    rmse(pred_gbm,   y)
  )
)


```
## Comparison visual

```{r visual_rmse}

df_compare <- data.frame(
  Truth = y,
  Stump = pred_stump,
  RF    = pred_rf,
  GBM   = pred_gbm
)

df_long <- pivot_longer(
  df_compare,
  cols = c("Stump", "RF", "GBM"),
  names_to = "Model",
  values_to = "Pred"
)

ggplot(df_long, aes(Truth, Pred)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~ Model) +
  theme_minimal() +
  labs(
    title = "Prediction vs Truth for Stump, Random Forest, and GBM",
    subtitle = "Perfect predictions lie on the dashed diagonal line",
    x = "True Values",
    y = "Predicted Values"
  )

```

## Visuals for synthetic regression dataset.

```{r visual_data1}

ggplot(df, aes(X1, y, color = X2)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient2(low = "blue", high = "red", mid = "white") +
  theme_minimal() +
  labs(
    title = "Synthetic Regression Data",
    subtitle = "Two regions defined by the sign of X1",
    x = "X1",
    y = "Target y",
    color = "X2"
  )

```

```{r visual_data2}
ggplot(df, aes(X1, X2, color = y)) +
  geom_point(alpha = 0.8) +
  scale_color_gradient2(low = "blue", high = "red", mid = "white") +
  theme_minimal() +
  labs(
    title = "Nonlinear Structure of the Synthetic Dataset",
    subtitle = "Clear boundary at X1 = 0",
    x = "X1",
    y = "X2",
    color = "y"
  )

```
